Tropical mathematics and the Lambda-calculus I

### 2023-07-31 CSL24/sectionLaw.tex
commit

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}, that it's worth recalling.

@@

On the one hand we have $\Lawv$-modules: these are triples $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.

@@

%\subsection{$\Lawv$-categories}
On the other hand we have Lawvere's \emph{generalized metric spaces}:
Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying 
$0  \geq X(x,x)$ and $X(y,z)+X(x,y)\geq  X(x,z)$. This structure is often referred to as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}. %, or as a \emph{quasi-metric space}. 
 A metric space, in the usual sense, is thus the same as a skeletal (i.e.~$X(x,y)=0$ implies $x=y$) and symmetric (i.e.~$X(x,y)=X(y,x)$) $\Lawv$-category.
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, as functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.

A $\Lawv$-category is \emph{complete} when all weighted colimits (cf.~Appendix for a precise definition) exist in $X$. This categorical notion of completeness can be seen as a generalization of the usual notion of Cauchy-completeness.
We let $\GMet$ indicate the category of complete $\Lawv$-categories and continuous (i.e.~colimit-preserving) $\Lawv$-enriched functors.

Now, it turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
$M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$. Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories. Conversely, complete $\Lawv$-categories are \emph{tensored}, that is, admit a $\Lawv$-module structure $\epsilon \otimes x$ (see the Appendix).
Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
%
%
%To conclude our correspondence between $\Lawv$-modules and complete $\Lawv$-categories, it remains to observe that the 
%two constructions leading from one structure to the other are one the inverse of the other: for any $\Lawv$-module $(M,\preceq,\star)$,
%$x\preceq_{M}y$ iff $M(y,x)=0$ iff $x\preceq y=0\star y$, and, from  
%$M(\epsilon \star x, y)= M(x,y)\dotdiv \epsilon$, we deduce $\epsilon\otimes x=\epsilon \star x$. 
%Conversely, 
%for any complete $\Lawv$-category $X$ and $x,y\in X$, one can check that 
%$X(y,x)=\inf\{ \epsilon \mid X(\epsilon\otimes y,x )=0\}$.
%
%This leads to the following:
%
%
%\begin{theorem}[cf.~\cite{Stubbe2006}]
%$\Mod$ and $\GMet$ are isomorphic categories.
%\end{theorem}

@@

To conclude, we show how the correspondence $\Mod\simeq \GMet$ lifts to a model of the differential $\lambda$-calculus, extending the co-Kleisli category $\LREL_{!}$.

First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting suitable properties of $\Mod$, we may define $!$ as the product of the symmetric algebras.  
This ensures that the coKleisli category $\Mod_{!}$ is a CCC.
Moreover, the constructions for $\Mod$ generalize those of $\LREL$, in the sense that $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$ and that $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.

Finally, general results {\color{red}CITARE} allow us to conclude 
\begin{theorem}
$\Mod_{!}\simeq\GMet_{!}$ is a CC$\partial$C, when equipped with 
the differential operator $E$ defined, for $f:!M\to N$, by 
$Ef(\alpha)=
\bigvee\left\{
f(\beta\cup [x]) \ \Big \vert  \ 
\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
\right\}
$, where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}.
\end{theorem}

>>>>

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}, that it's certainly worth recalling here, albeit very briefly (more details in the Appendix).

@@

On the one hand we have $\Lawv$-modules: these are triples $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$. % (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$). 
A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.

@@

%\subsection{$\Lawv$-categories}
On the other hand we have Lawvere's \emph{generalized metric spaces}  \cite{Lawvere1973, Hofmann2014, Stubbe2014}:
Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying 
$0  \geq X(x,x)$ and $X(y,z)+X(x,y)\geq  X(x,z)$. 
This structure clearly generalizes the usual definition of metric spaces, which indeed are the same as skeletal (i.e.~$X(x,y)=0$ implies $x=y$) and symmetric (i.e.~$X(x,y)=X(y,x)$) $\Lawv$-categories.
Moreover, a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is nothing but a non-expansive map $f:X\to Y$, since functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.

Lawvere also observed that the usual notion of Cauchy-completeness can be formulated, in this framework, as the existence of suitable colimits \cite{Lawvere1973}. A $\Lawv$-category is said \emph{complete} when \emph{all} {weighted colimits} exist in $X$. Let $\GMet$ indicate the category of complete $\Lawv$-categories and continuous (i.e.~colimit-preserving) $\Lawv$-enriched functors.

It turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
$M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$. Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories. Conversely, complete $\Lawv$-categories are \emph{tensored} (cf.~\cite{Stubbe2014}), and thus admit a suitable $\Lawv$-module structure $\epsilon \otimes x$.
Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
%
%
%To conclude our correspondence between $\Lawv$-modules and complete $\Lawv$-categories, it remains to observe that the 
%two constructions leading from one structure to the other are one the inverse of the other: for any $\Lawv$-module $(M,\preceq,\star)$,
%$x\preceq_{M}y$ iff $M(y,x)=0$ iff $x\preceq y=0\star y$, and, from  
%$M(\epsilon \star x, y)= M(x,y)\dotdiv \epsilon$, we deduce $\epsilon\otimes x=\epsilon \star x$. 
%Conversely, 
%for any complete $\Lawv$-category $X$ and $x,y\in X$, one can check that 
%$X(y,x)=\inf\{ \epsilon \mid X(\epsilon\otimes y,x )=0\}$.
%
%This leads to the following:
%
%
%\begin{theorem}[cf.~\cite{Stubbe2006}]
%$\Mod$ and $\GMet$ are isomorphic categories.
%\end{theorem}

@@

To relate this elegant correspondence with our work on $\LREL_{!}$, we show that it lifts to a model of the differential $\lambda$-calculus, extending the co-Kleisli category $\LREL_{!}$. 
First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting suitable properties of $\Mod$, we may define $!$ as the product of the symmetric algebras.  
This ensures that the coKleisli category $\Mod_{!}$ is a CCC.
Moreover, the constructions for $\Mod$ generalize those of $\LREL$, in the sense that $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$ and that $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.
Finally, as the coKleisli category of a Lafont category with biproducts is always a CC$\partial$C \cite{LemayCALCO2021}, we can endow $\Mod_{!}$ with a suitable differential operator $E$ generalizing $D^{!}$ (see the Appendix) so that:
\begin{theorem}
($\Mod_{!}/\GMet_{!},E$) is a CC$\partial$C.
%, when equipped with 
%the differential operator $E$ defined, for $f:\ !M\to N$, by 
%$Ef(\alpha)=
%\bigvee\left\{
%f(\beta\cup [x]) \ \Big \vert  \ 
%\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
%\right\}
%$, where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
%and $S: \ !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}.
\end{theorem}

### 2023-07-31 CSL24/sectionTaylip.tex
Commit dettagli
commit

We now have all elements necessary to relate the metric and differential analysis of higher-order programs in the tropical relational model.

The key ingredient is the notion of Taylor expansion $\Te{M}$ of a $\lambda$-term $M$, which is a set of terms of the differential $\lambda$-calculus defined as follows:
%
%
%
% which is defined as follows: 
%
%that we now quickly recall. 
%
%
%Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
%$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
%%
%
%The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as 
%
$\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
$\Te{MN}=\{t\cdot \langle u_{1},\dots, u_{k}\rangle  \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$, where $t\cdot \langle u_{1},\dots, u_{k}\rangle$ is an abbreviation for $\mathsf D^{k}[t, u_{1},\dots, u_{k}] 0$. 
Observe that in the terms appearing in $\Te{M}$ all applications are always \emph{bounded}: they may use an exact number of copies of their input. 
Such terms are usually called \emph{resource $\lambda$-terms} \cite{Pagani2009, Manzo2012}.
Considering the term $M=zxx$ from Example \ref{ex:zxx}, all terms
 $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tps. 
However, resource $\lambda$-terms need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
with $t=y\cdot \langle y\cdot \langle x\rangle \rangle\in \Te{y(yx)}$. Then 
$\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
\times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
$
\model{t}^{!}(y,x)_{i}= \inf_{m,n\in \mathbb N}\big\{    y_{[m],i}  +  y_{[n],m}+  x_{n}\}
$, 
which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown below.

%
%$\Te{MN}$ of $MN$ is the set $\set{\Der^{n}[t,u_1,\dots,u_n]0 \mid n\in\N,t\in\Te{M},u_i\in\Te{N}}$.
We have already shown that the tropical differential makes $\LREL_{!}$ a model of the differential $\lambda$-calculus. We now show that it also models the Taylor expansion (this needs not be true for \emph{any} CC$\partial$C).
First, it can be patiently checked that in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):

\begin{theorem}\label{thm:modelsTaylor}
 For all $t\in\HOM{\LREL_!}{Z}{!X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align} 
\end{theorem}
%It is worth discussing the formula above a bit more. 
The equation above is a nothing more than a tropical version of the Taylor formula from the Introduction:
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$.
%,  i.e.~it coincides with the tropical %interpretation of the
%version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.

@@

Using the results of the previous section, as well as the results above, we can deduce the following properties:

@@

\item For $\mathcal S=\STDLC$, if $M\in \Te{M'}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is Lipschitz;

@@

\end{theorem}
\begin{proof}
1). As we already observed, the interpretation of a bounded term is a tropical polynomial.
2.) From Proposition \ref{prop:troplinear} 2.~observing that a resource term $t(x)$ may use a variable $x$ a fixed number $n$ times, so that its matrix lies in $\Lawv^{!_{n}X\times Y}$. 
%Now we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and by taking the maximum Lipschitz constant among the finite number $\mathrm{Card}(\model A)$ of them, we obtain the thesis.
3). From Theorem~\ref{thmTLSlocLip}.
4). It follows from \autoref{cor:T(M)=M} plus the fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
\end{proof}

@@

\begin{example}
Consider again the term $M=zxx$ from Example \ref{ex:zxx}-
%, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 
%
% $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
 The (generalized) tps $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
$L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$. 
Now take the approximant $t= z \langle x^{N-1}\rangle \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
$\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= 3N+3 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
\end{example}
%    
%    %
%  Consider now $F\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ given by $F_{\mu}=0$ if $\mu=[(1,1)]$ and $F_{\mu}=\infty$ otherwise. 
%  Then $\model{M}^{!}(\langle x,F\rangle)=2x$
%%We show that we can compute a local 
%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:

>>>>

In this section we finally relate the metric and differential analysis of higher-order programs in the tropical relational model.

The key ingredient is the notion of Taylor expansion $\Te{M}$ of a $\lambda$-term $M$. This is a set of terms of the differential $\lambda$-calculus defined inductively as: 
%
%
%
% which is defined as follows: 
%
%that we now quickly recall. 
%
%
%Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
%$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
%%
%
%The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as 
$\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
$\Te{MN}=\{t\cdot \langle u_{1},\dots, u_{k}\rangle  \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$, where $t\cdot \langle u_{1},\dots, u_{k}\rangle$ is an abbreviation for $\mathsf D^{k}[t, u_{1},\dots, u_{k}] 0$. 
Observe that in the terms appearing in $\Te{M}$ all applications are \emph{bounded}: they may use an exact number of copies of their input. 
Such terms are usually called \emph{resource $\lambda$-terms} \cite{Pagani2009, Manzo2012}. One can easily check that for all terms $\Gamma \vdash_{\STLC}M:A$ and  $t\in \Te{M}$, also 
$\Gamma \vdash_{\STDLC}t:A$ holds. 
\begin{example}
Considering the term $M=zxx$ from Example \ref{ex:zxx}, all terms
 $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
Notice that the interpretation of $t_{n,m}$ yields a tropical polynomial
$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tps. 
However, this is not a general fact: consider $y:(o\to o )\to (o\to o), x:(o\to o) \vdash t:  (o\to  o)$
with $t=y\cdot \langle y\cdot \langle x\rangle \rangle\in \Te{y(yx)}$. Then 
$\model{t}^{!}: \Lawv^{  !\mathbb N\times \mathbb N}
\times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
$
\model{t}^{!}(y,x)_{i}= \inf_{m,n\in \mathbb N}\big\{    y_{[m],i}  +  y_{[n],m}+  x_{n}\}
$, 
which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown by Theorem \ref{thm:taylor} below.
\end{example}

%
%$\Te{MN}$ of $MN$ is the set $\set{\Der^{n}[t,u_1,\dots,u_n]0 \mid n\in\N,t\in\Te{M},u_i\in\Te{N}}$.
We have already shown that the tropical differential makes $\LREL_{!}$ a model of the differential $\lambda$-calculus. We now show that it also models the Taylor expansion (this needs not be true for \emph{any} CC$\partial\lambda$C).
First, it can be patiently checked that (see \cite[Definition 4.22]{Manzo2012}):
%
% in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):
\begin{theorem}\label{thm:modelsTaylor}
Morphisms in $(\LREL_{!},D)$ can be Taylor-expanded: for all $t\in\HOM{\LREL_!}{Z}{!X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have %, the evaluation of $t$ over $s$ yields 
% \begin{align}\label{eq:taylorcat}
  $\RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.$
% \end{align} 
\end{theorem}
%It is worth discussing the formula above a bit more. 
The equation above is a tropical reformulation of the Taylor formula from the Introduction:
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$.
%,  i.e.~it coincides with the tropical %interpretation of the
%version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
%Moreover, one can show that the metrics $d_{\infty}$ are compatible with the Taylor formula in the following sense:
%\begin{proposition}
%For all tps $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
%the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
%$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
%where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
%%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
%%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
%\end{proposition}

@@

Using the results of the previous section, as well as the results above, we now deduce the following properties:

@@

\item For $\mathcal S=\STDLC$, if $t\in \Te{M}$, then $\model{\Gamma\vdash_{\mathcal S}t:A}^{!}_{a}$ is Lipschitz;

@@

\end{theorem}
\begin{proof}
1). From Proposition~\ref{prop:troplinear} 2.~and the remark that for any type of $\BSTLC$, $\model{A}$ is finite.
2.) From Proposition~\ref{prop:troplinear} 2.~observing that a resource term $t(x)$ may use a variable $x$ a fixed number $n$ times, so that its matrix lies in $\Lawv^{!_{n}X\times Y}$. 
%Now we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and by taking the maximum Lipschitz constant among the finite number $\mathrm{Card}(\model A)$ of them, we obtain the thesis.
3). From Theorem~\ref{thmTLSlocLip}.
4). It follows from \autoref{cor:T(M)=M} plus the fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have 
%$\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with 
$\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
\end{proof}

@@

\begin{example}
Consider again the term $M=zxx$ from Example \ref{ex:zxx}. 
%, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 
%
% $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
 The (generalized) tps $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
$L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$. 
Now take the approximant $t= z \langle x^{N-1}\rangle \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
$\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= 3N+3 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
\end{example}
%    
%    %
%  Consider now $F\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ given by $F_{\mu}=0$ if $\mu=[(1,1)]$ and $F_{\mu}=\infty$ otherwise. 
%  Then $\model{M}^{!}(\langle x,F\rangle)=2x$
%%We show that we can compute a local 
%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
